Timbral similarity between extended playing techniques: the next frontier in querying instrumental databases

### 2018-01-29 paper/lagrangeSpontaneousSimilarity2017.tex
volountary -> voluntary
fix spelling errors

The plane is displayed on a computer screen using a web based interface. The \ipts can be listened to by hovering the mouse on the corresponding dot. The experiment is over when the sujbect assigned a color to all the dots. The initial location of the dots on the plane is randomized for each subject and can be changed during the experiment by the subject for convenience. Eventhough the location of the dots is recorded, this data is not considered during the analysis.

The experiment was available online for two months and requests to perform the test have been sent to internal mailing lists of the CNSMDP and international research mailing with a focus on audio and music processing. The experiment have been performed on a volountary basis without gratification. Subjects were asked to use headphones at a comfortable listening level.

@@

As discussed in the previous section, only the clustering given by the subjects using the color labels is considered to estimate the spontaneous judgments of similarity among \ipt. The spatial organization of the dots representing the \ipt could provide information about the similarity, but this would implicitely force the fact that the timbral similarity space is two dimensional. We therefore now only study the properties of the clusterings performed by the subject prior to converting them into an overall similarity matrix.

@@

A clustering can be considered as a binary similarity measure, with $s(a, b) = 1$ if $a$ and $b$ belong to the same group, and $0$ otherwise. By averaging the 31 binary similarity matrix, an overall floating point matrix is obtained that correponds to our measure of the spontaneous perceptual similarity between \ipt.

@@

There is a clear gain of NMI when considering the playing technique which means that the subjects primarily used cues corresponding to the playing technique to organize the \ipts. Also, for both curves, a global trend of increase in the $j/x$ partitions with respect to the number of clusters can be observed. This is a well documented fact, which simply put, means that it is easier to match any given partition with more clusters at hand \cite{tibshirani2001estimating}. To compensate for this bias, one can consider the NMI that would be achieved by comparing to the reference partition several random partitions of increasing number of clusters. The resulting averaged NMI for 100 randomly generated partitions is displayed on Figure \ref{fig:clusters}. This so-called null partitions NMI are respectively termed $nullI$ and $nullPt$ depending on the reference used. To normalize $j/I$ and $j/Pt$ in order to reduce the impact of this bias, we simply substract the $null$ NMI:
\begin{eqnarray}
  nj/I &=& j/I - null \\
  nj/Pt &=& j/Pt - null  \\
\end{eqnarray}
This normalization is useful for identifying more easily the number of clusters for which the comparison is most relevant. As can be seen on Figure \ref{fig:clusters}, the number of clusters that leads to the highest NMI is 5 for the playing technique reference and 8 for the instrument reference, leading respectively to an NMI of $0.46$ and $0.21$. For the latter, reducing the number of clusters down to 6 do not decrease the NMI by much, a fact that will be latter discussed in more details. Even by considering the null normalization, the matching level is much higher for the playing technique reference.

@@

Though, even with an optimal number of clusters, we see that the clusters have  a low purity, \ie{} a very diverse set of instruments labels for the \ipt they belong to. Also, some clusters could be merged without stong changes in the overall organization. This probably explains the small changes of NMI when reducing the number of clusters, see Figure \ref{fig:clusters}.

@@

Once the perceptual data is gathered and controled, our aim is now to evaluate the ability of the proposed two stages perceptual model brifely described and motivated in the Introduction. The model is composed of two successive processing stages, respectively called the primary and the secondary stage, see Figure \ref{fig:model}. The primary stage is designed to rely on very few meta parameters set arbitrarily by the experimenter to mimic the early stages of the mamalian auditory system. On contrary, the secondary stage rely largely on the data it has to predict and in this respect mimic the high plasticity of the later stages of the mamalian auditory system.

@@

In order to model the early stages of the mamalian auditory system, we use the time scattering transform (TST). This transform is obtained by applying a one dimensional
wavelet transform in time to a time-frequency wavelet scalogram \cite{anden2014deep}. It is thus well suited to encode modulations over time of energy distributed over frequency, dimensions that are shown to be perceptually important \cite{dau1997modeling}.

@@

The secondary stage involves a learning based process block that will weight and linearly combine the features outputed by the primary stage.

@@

Another solution is to compute a projection matrix that maximizes the fact that neighbors in the representational space belong to the same cluster in the reference clustering while items from different clusters are separated by a large margin. To do so, the large margin nearest neighbor (LMNN) metric learning algorithm \cite{weinberger2006distance, weinberger2009distance} is considered. One advantage of the latter is that the resulting representational space is of the same dimensionality as the input one whereas the LDA output dimentionality is contrained by the number of classes in the reference clustering.

@@

Another approach is to combine the projection matrices issued from the processing of the projection algorithm over the 31 reference clusterings. As the number of classes can vary between reference clusterings, the kind of combination is not feasible using the LDA method. On contrary, the LMNN method aloways outputs a projection matrix with equal dimensions. Those projection matrices being of the same dimensionality they can safely be averaged to estimate an consensus projection.

@@

As it will be discussed in the next Section, the main application scenario that is considered in this paper is the recommandation of sounds based on examplar. This involves the definition of a similarity measure that behave similarly to the human perception. In this paper, the reference used for assessing closseness to human perception is the perceptual data gathered in experiment 2.

@@

As explained in the first section, gathering perceptual similarity data is plagued is plagued with a size problem. Thus, the perceptual data considers 78 \ipts each represented by one sound recording, most often of C4 pitch and nuance mezzo forte. In order to provide enough data for the numerical analysis described in this section, we complement the data gathered during the second experiment as follows: for each \ipt, instead of one sound examplar we consider the whole range of nuance and pitch that is available. This leads to a dataset of 9346 sound recordings. It is thus assumed that the perceptual similarity is invariant with respect to pitch and nuance which is not necessarily true, but is taken in this paper as a reasonable assumption.
\vl{It's too bad that this "reasonable" assumption is not challenged against the data. In music psychology, the "bandwidth for timbre invariance" is of the order of one octave (Handel and Erickson, Music Perception, 2001) or less (Steele and Williams, Music Perception, 2006).
One way to do it without collecting new similarity judgments would be to break down the precision by pitch of query.
Presumably we have an inverted U shape with a peak near C4.
@research @computation @figure @biblio}

@@

To evaluate the performance of the proposed approach, several baselines are considered for comparison purposes. For the first stage, the well known Mel-Frequency Cepstral Coefficients (MFCC)s are considered \cite{rabiner1993fundamentals}. The MFCCs are computed from audio frame of 25ms to 1 second, the number of mel bands is 40 and all the discrete cosine transform coefficients are kept. The resulting features have thus a dimensionality of 40. The MFCCs are then standardized by removing the mean and dividing by the standard deviation computed over the whole dataset.

@@

The LDA method do not requires any setting of parameters and the LMNN method is set to optimize for a neigbors size of 5.

@@

In this experiment, the consensus approach is favored for TST but not for MFCCs. This approach has the considerable advantage of being more computationaly efficicient as with only one reference to optimize from the LMMN training is much faster. Though, it is very likely that for a larger number of reference clusterings, the multiple LMNN would be useful to better model the diversity of the references.

@@

We investigated the capabilities of the proposed system to model a perceptual space averaged over several subjects. For application purposes that will be discussed in the next Section, it can be useful to consider the reference clustering of only one user and see if the system perform well with this kind of reference. To do so, the evaluation protocol is slightly modified. First the secondary stage is optimized for the 31 reference clusterings independantly and for each resulting projection the precision is computed. The 31 precision measures are then averaged and shown in Table \ref{tab:res2}. The excellent results of the TST optimized using LMNN demonstrates the usefullness of the proposed approach to model complex similarity space such as the percpetual ones gathered in Experiment 2.

@@

For the sake of reproducibility, the code of the proposed method and the experimental protocol is available\footnote{https://github.com/mathieulagrange/paperSpontaneousSimilarity}. For convenience, the features (scattering and mfccs) are also provided, but the software can be used to process audio data.

@@

An interesting alternative type of search is the similarity content based search. Being solely based on sound, this kind of search can be used when querying by example. The user provides an example of the target sound to find \ipt that match the timbre description of the query, be it an onomatepea, a tone from another instrument or any kind of sound.

The computational model proposed in this paper is well suited for this kind of search, as it can model complex perceptual timbre spaces and can scale to large databases. Once the features and the projection matrix are computed, the similarity computation is quite effective and can be further enhanced by considering fast similarity search scheme such as locally sensitive least square hashing \cite{pauleve2010locality}. The secondary stage being learnt, it can easily adapt to user tastes and usage history by reweighting the features for improved search relevance.

@@

The modeling of the spontaneous perceptual similarity have been studied. Experimental data have been collected using two perceptual tests whose outomes have been analysed to produce reference data of spontaneous perceptual judgments. This data is then modeled computationaly using a two step model.

@@

Computational experiments showed an interesting phenomenon that high dimensional features spaces such as the one achieved using the TST are most useful if another computation step is considered, if this second step is able to incorporate some sort of prior knowledge regarding the task at hand in order to refine the resulting representation. It would be interesting to verify if this phenomenon can be observed for other types of targeted tasks, using larger datasets and with a higher level of control regarding the learning procedure to study the impact of potential overfiting or learning biases of this king of techniques.

>>>>

The plane is displayed on a computer screen using a web based interface. The \ipts can be listened to by hovering the mouse on the corresponding dot. The experiment is over when the subject assigned a color to all the dots. The initial location of the dots on the plane is randomized for each subject and can be changed during the experiment by the subject for convenience. Event hough the location of the dots is recorded, this data is not considered during the analysis.

The experiment was available online for two months and requests to perform the test have been sent to internal mailing lists of the CNSMDP and international research mailing with a focus on audio and music processing. The experiment have been performed on a voluntary basis without gratification. Subjects were asked to use headphones at a comfortable listening level.

@@

As discussed in the previous section, only the clustering given by the subjects using the color labels is considered to estimate the spontaneous judgments of similarity among \ipt. The spatial organization of the dots representing the \ipt could provide information about the similarity, but this would implicitly force the fact that the timbral similarity space is two dimensional. We therefore now only study the properties of the clusterings performed by the subject prior to converting them into an overall similarity matrix.

@@

A clustering can be considered as a binary similarity measure, with $s(a, b) = 1$ if $a$ and $b$ belong to the same group, and $0$ otherwise. By averaging the 31 binary similarity matrix, an overall floating point matrix is obtained that corresponds to our measure of the spontaneous perceptual similarity between \ipt.

@@

There is a clear gain of NMI when considering the playing technique which means that the subjects primarily used cues corresponding to the playing technique to organize the \ipts. Also, for both curves, a global trend of increase in the $j/x$ partitions with respect to the number of clusters can be observed. This is a well documented fact, which simply put, means that it is easier to match any given partition with more clusters at hand \cite{tibshirani2001estimating}. To compensate for this bias, one can consider the NMI that would be achieved by comparing to the reference partition several random partitions of increasing number of clusters. The resulting averaged NMI for 100 randomly generated partitions is displayed on Figure \ref{fig:clusters}. This so-called null partitions NMI are respectively termed $nullI$ and $nullPt$ depending on the reference used. To normalize $j/I$ and $j/Pt$ in order to reduce the impact of this bias, we simply subtract the $null$ NMI:
\begin{eqnarray}
  nj/I &=& j/I - null \\
  nj/Pt &=& j/Pt - null  \\
\end{eqnarray}
This normalization is useful for identifying more easily the number of clusters for which the comparison is most relevant. As can be seen on Figure \ref{fig:clusters}, the number of clusters that leads to the highest NMI is 5 for the playing technique reference and 8 for the instrument reference, leading respectively to an NMI of $0.46$ and $0.21$. For the latter, reducing the number of clusters down to 6 do not decrease the NMI by much, a fact that will be latter discussed in more details. Even by considering the null normalization, the matching level is much higher for the playing technique reference.

@@

Though, even with an optimal number of clusters, we see that the clusters have  a low purity, \ie{} a very diverse set of instruments labels for the \ipt they belong to. Also, some clusters could be merged without strong changes in the overall organization. This probably explains the small changes of NMI when reducing the number of clusters, see Figure \ref{fig:clusters}.

@@

Once the perceptual data is gathered and controlled, our aim is now to evaluate the ability of the proposed two stages perceptual model brifely described and motivated in the Introduction. The model is composed of two successive processing stages, respectively called the primary and the secondary stage, see Figure \ref{fig:model}. The primary stage is designed to rely on very few meta parameters set arbitrarily by the experimenter to mimic the early stages of the mamalian auditory system. On contrary, the secondary stage rely largely on the data it has to predict and in this respect mimic the high plasticity of the later stages of the mamalian auditory system.

@@

In order to model the early stages of the mammalian auditory system, we use the time scattering transform (TST). This transform is obtained by applying a one dimensional
wavelet transform in time to a time-frequency wavelet scalogram \cite{anden2014deep}. It is thus well suited to encode modulations over time of energy distributed over frequency, dimensions that are shown to be perceptually important \cite{dau1997modeling}.

@@

The secondary stage involves a learning based process block that will weight and linearly combine the features produced by the primary stage.

@@

Another solution is to compute a projection matrix that maximizes the fact that neighbors in the representational space belong to the same cluster in the reference clustering while items from different clusters are separated by a large margin. To do so, the large margin nearest neighbor (LMNN) metric learning algorithm \cite{weinberger2006distance, weinberger2009distance} is considered. One advantage of the latter is that the resulting representational space is of the same dimensionality as the input one whereas the LDA output dimensionality is constrained by the number of classes in the reference clustering.

@@

Another approach is to combine the projection matrices issued from the processing of the projection algorithm over the 31 reference clusterings. As the number of classes can vary between reference clusterings, the kind of combination is not feasible using the LDA method. On contrary, the LMNN method always outputs a projection matrix with equal dimensions. Those projection matrices being of the same dimensionality they can safely be averaged to estimate an consensus projection.

@@

As it will be discussed in the next Section, the main application scenario that is considered in this paper is the recommendation of sounds based on exemplar. This involves the definition of a similarity measure that behave similarly to the human perception. In this paper, the reference used for assessing similarity to human perception is the perceptual data gathered in experiment 2.

@@

As explained in the first section, gathering perceptual similarity data is plagued is plagued with a size problem. Thus, the perceptual data considers 78 \ipts each represented by one sound recording, most often of C4 pitch and nuance mezzo forte. In order to provide enough data for the numerical analysis described in this section, we complement the data gathered during the second experiment as follows: for each \ipt, instead of one sound exemplar we consider the whole range of nuance and pitch that is available. This leads to a dataset of 9346 sound recordings. It is thus assumed that the perceptual similarity is invariant with respect to pitch and nuance which is not necessarily true, but is taken in this paper as a reasonable assumption.
\vl{It's too bad that this "reasonable" assumption is not challenged against the data. In music psychology, the "bandwidth for timbre invariance" is of the order of one octave (Handel and Erickson, Music Perception, 2001) or less (Steele and Williams, Music Perception, 2006).
One way to do it without collecting new similarity judgments would be to break down the precision by pitch of query.
Presumably we have an inverted U shape with a peak near C4.
@research @computation @figure @biblio}

@@

To evaluate the performance of the proposed approach, several baselines are considered for comparison purposes. For the first stage, the well-known Mel-Frequency Cepstral Coefficients (MFCC)s are considered \cite{rabiner1993fundamentals}. The MFCCs are computed from audio frame of 25ms to 1 second, the number of mel bands is 40 and all the discrete cosine transform coefficients are kept. The resulting features have thus a dimensionality of 40. The MFCCs are then standardized by removing the mean and dividing by the standard deviation computed over the whole dataset.

@@

The LDA method do not requires any setting of parameters and the LMNN method is set to optimize for a neighbors size of 5.

@@

In this experiment, the consensus approach is favored for TST but not for MFCCs. This approach has the considerable advantage of being more computationally efficient as with only one reference to optimize from the LMMN training is much faster. Though, it is very likely that for a larger number of reference clusterings, the multiple LMNN would be useful to better model the diversity of the references.

@@

We investigated the capabilities of the proposed system to model a perceptual space averaged over several subjects. For application purposes that will be discussed in the next Section, it can be useful to consider the reference clustering of only one user and see if the system perform well with this kind of reference. To do so, the evaluation protocol is slightly modified. First the secondary stage is optimized for the 31 reference clusterings independently and for each resulting projection the precision is computed. The 31 precision measures are then averaged and shown in Table \ref{tab:res2}. The excellent results of the TST optimized using LMNN demonstrates the usefulness of the proposed approach to model complex similarity space such as the perceptual ones gathered in Experiment 2.

@@

For the sake of reproducibility, the code of the proposed method and the experimental protocol is available\footnote{https://github.com/mathieulagrange/paperSpontaneousSimilarity}. For convenience, the features (scattering and MFCCs) are also provided, but the software can be used to process audio data.

@@

An interesting alternative type of search is the similarity content based search. Being solely based on sound, this kind of search can be used when querying by example. The user provides an example of the target sound to find \ipt that match the timbre description of the query, be it an onomatopoeia, a tone from another instrument or any kind of sound.

The computational model proposed in this paper is well suited for this kind of search, as it can model complex perceptual timbre spaces and can scale to large databases. Once the features and the projection matrix are computed, the similarity computation is quite effective and can be further enhanced by considering fast similarity search scheme such as locally sensitive least square hashing \cite{pauleve2010locality}. The secondary stage being learnt, it can easily adapt to user tastes and usage history by re-weighting the features for improved search relevance.

@@

The modeling of the spontaneous perceptual similarity have been studied. Experimental data have been collected using two perceptual tests whose outcomes have been analyzed to produce reference data of spontaneous perceptual judgments. This data is then modeled computationally using a two step model.

@@

Computational experiments showed an interesting phenomenon that high dimensional features spaces such as the one achieved using the TST are most useful if another computation step is considered, if this second step is able to incorporate some sort of prior knowledge regarding the task at hand in order to refine the resulting representation. It would be interesting to verify if this phenomenon can be observed for other types of targeted tasks, using larger datasets and with a higher level of control regarding the learning procedure to study the impact of potential overfitting or learning biases of this king of techniques.
